A Comparision of Bootstrap Confidence Interval Methods

### 2022-12-09 project/PaperTemplate/paper.Rmd
Minor edits

abstract: |
  This paper explores different bootstrap methods for creating confidence intervals and tests their performance against each other for different types of distributions. This paper will re-introduce the ideas of bootstraps and confidence intervals, as well as the theory behind different methods of bootstrapped confidence intervals. Along with the theoretical details of each method, we will provide an example of each bootstrapped confidence interval method with a toy dataset. Then, we will run a simulation that evaluates the performance of a couple of the methods we introduced on a variety of sample sizes and types of distributions.

@@

We demonstrate performing a non-parametric bootstrap below using `SnowGR`, which gives the official snowfall dataset by the month for Grand Rapids, MI, starting in 1893. 
We will be using the `Dec` variable, which is the number of inches of snow that fell in December of each year. 
In particular, we are interested in the sample mean and later finding confidence intervals for the population mean of `Dec`. 
The distribution is shown (Figure 1).
```{r SnowGR histogram}
data(SnowGR)
gf_histogram(~ Dec, data = SnowGR) +
  labs(x = "Annual Snowfall in December", 
       title = "Distribution of Annual Snowfall in December", y = "Count",
       caption = "Fig. 1: Histogram of SnowGR")
```
Next, we will perform the non-parametric bootstrap using 10,000 replications.
```{r non-paramteric bootstrap}
set.seed(495)
orig_mean <- mean(~ Dec, data = SnowGR) # theta hat

set.seed(495)
nboot <- 10000

mean <- rep(0,nboot)
sd <- rep(0,nboot)
se <- rep(0,nboot)
t_stat <- rep(0,nboot)

for (i in 1:nboot) {
  resampled <- as.data.frame(mosaic::resample(SnowGR, replace = TRUE))
  mean[i] <- mean(~Dec, data = resampled)
  sd[i] <- sd(~Dec, data = resampled)
}

dec_means <- as.data.frame(mean)
```
Figure 2 shows the histogram of bootstrapped values. Since we performed a large number of bootstrap replications, the bootstrapped distribution appears to be approximately Normal.
```{r, cache = TRUE}
gf_dhistogram(~ mean, data = dec_means) %>%
  gf_dens() %>%
  gf_labs(title = "Histogram of 10,000 Bootstrapped Mean Dec Values",
          caption = "Fig. 2: Bootstrap histogram of Dec values")
```

@@

```{r}
gf_histogram(~ Dec, data = SnowGR) +
  labs(x = "Annual Snowfall in December", 
       title = "Distribution of Annual Snowfall in December", y = "Count",
       caption = "Fig. 3: Histogram with Standard Interval Shown") +
  geom_vline(aes(xintercept= 13.85639)) +
  geom_vline(aes(xintercept= 17.65957))
```

@@

```{r}
qdata(~ mean, c(0.025, 0.975), data = dec_means)
```
Therefore, we can say that we are 95% confident that the true population mean of inches of snow that fall in December is between 13.91 inches and 17.72 inches. Our original histogram with the percentile confidence interval is shown (Figure 4).
```{r}
gf_histogram(~ Dec, data = SnowGR) +
  labs(x = "Annual Snowfall in December", 
       title = "Distribution of Annual Snowfall in December", y = "Count",
       caption = "Fig. 4: Histogram with Percentile Interval Shown") +
  geom_vline(aes(xintercept= 13.91170)) +
  geom_vline(aes(xintercept= 17.72353))
```
This interval is similar to the one created using the standard method, but shifted slightly to the right.

@@

```{r}
gf_histogram(~ Dec, data = SnowGR) +
  labs(x = "Annual Snowfall in December", 
       title = "Distribution of Annual Snowfall in December", y = "Count",
       caption = "Fig. 5: Histogram with BC Interval Shown") +
  geom_vline(aes(xintercept= 14.01172)) +
  geom_vline(aes(xintercept= 17.82017))

@@

```{r}
gf_histogram(~ Dec, data = SnowGR) +
  labs(x = "Annual Snowfall in December", 
       title = "Distribution of Annual Snowfall in December", y = "Count",
       caption = "Fig. 6: Histogram with BCa Interval Shown") +
  geom_vline(aes(xintercept= 13.91591)) +
  geom_vline(aes(xintercept= 17.72353))

@@

```{r}
gf_histogram(~ Dec, data = SnowGR) +
  labs(x = "Annual Snowfall in December", 
       title = "Distribution of Annual Snowfall in December", y = "Count", 
       caption = "Fig. 7: Histogram with Studentized Interval Shown") +
  geom_vline(aes(xintercept= 14.03467)) +
  geom_vline(aes(xintercept= 18.01709)) 
```

@@

## Code
For our simulation, we will randomly sample from distributions in which we know the true population mean. 
For example, we will be sampling from a Gamma distribution where we can find the true population mean using the shape and rate parameters of the sample distribution. 
After we sample from the distribution, we will create a bootstrapped confidence interval for a few of the methods we described in the exposition. 
Since we know the true population mean of the distribution, we will determine if the true population mean is contained in the confidence interval for each of the methods we are testing [@Guillaume21]. 
We will perform this process a large number of times (10,000 simulations) and then average by the number of iterations to determine, on average, the proportion that each bootstrapped confidence interval method contains the true population mean for a given distribution. 
We will do this for a couple of different distributions and ranges of sample sizes to show how the intervals perform with skewed distributions and larger sample sizes.

To do this simulation, we will use the `boot` package [@boot]. 
This package can compute a bootstrap object which contains the output from a bootstrap calculation of a given sample and then produces confidence intervals from the bootstrap object. 
In particular, we will use the `boot.ci` function to create four confidence intervals: the standard (or normal) confidence interval, the studentized confidence interval, the percentile confidence interval, and the bias-corrected with acceleration confidence interval.

@@

```{r normal simulation}
run_sim_normal <- function(sim_reps) {
  # create function to return mean and sd of bootstrap replication
  boot.stat <- function(d, i) {
    d <- X[i]
    return (c(mean(d), sd(d)))
  }
  # initialize matrix to hold different bootstrap prop. for each sample size
  master <- matrix(nrow = num_sample_sizes, ncol = 5)
  # iterate through each sample size
  for(i in 1:num_sample_sizes) {
    # initialize vectors that hold if the CI contains true mean by iteration
    norm_contains <- rep(0, sim_reps)
    student_contains <- rep(0, sim_reps)
    percentile_contains <- rep(0, sim_reps)
    bca_contains <- rep(0, sim_reps)
    # 
    n <- sample_sizes[i]
    true_mean <- 0
    # perform 10,000 iterations of bootstraps
    for (j in 1:sim_reps) {
      X <- rnorm(n, 0, 1)
      values <- data.frame(X)
      # perform bootstrap and CIs
      results <- boot(data = values, statistic = boot.stat, R = boot_reps)
      ci <- boot.ci(results, conf = 0.95, 
                    type = c("norm", "perc", "bca", "stud"))
      # update vectors depending on if the CI contains true mean
      norm_contains[j] <- true_mean >= ci$normal[2] & 
        true_mean <= ci$normal[3]
      student_contains[j] <- true_mean >= ci$student[4] & 
        true_mean <= ci$student[5]
      percentile_contains[j] <- true_mean >= ci$percent[4] & 
        true_mean <= ci$percent[5]
      bca_contains[j] <- true_mean >= ci$bca[4] &
        true_mean <= ci$bca[5]
    }
    # add the sample size and average proportion to matrix
    master[i, 1] <- sample_sizes[i]
    master[i, 2] <- sum(norm_contains)/sim_reps
    master[i, 3] <- sum(student_contains)/sim_reps
    master[i, 4] <- sum(percentile_contains)/sim_reps
    master[i, 5] <- sum(bca_contains)/sim_reps
  }
  # some wrangling
  master_df <- as.data.frame(master) %>% 
      rename("Sample_Size" = V1, "Normal" = V2, "Studentized" = V3, 
             "Percentile" = V4, "BCa" = V5) 
  master_df
}
```

@@

normal_fig <- ggplot(data = normal_df_long, 
                    aes(x = Sample_Size, y = prop, color = method)) + 
  geom_point() + geom_line() + 
  labs(
    title = "Prop. of CIs that Contain the True Mean from Normal Distibution", 
    x = "Sample Size", 
    y = "Proportion (of 10,000 Simulations)", color = "Method",
    caption = "Fig. 8: Graph showing the Accuracy of Bootstrap CI Methods for
       different sample sizes on Normal Dist.") +
  theme_light() +
  theme(legend.position="bottom")

@@

gamma_fig <- ggplot(data = gamma_df_long, 
                    aes(x = Sample_Size, y = prop, color = method)) + 
  geom_point() + geom_line() + 
  labs(title = "Prop. of CIs that Contain the True Mean from Gamma Distibution", 
       x = "Sample Size", 
       y = "Proportion (of 10,000 Simulations)", color = "Method",
       caption = "Fig. 9: Graph showing the Accuracy of Bootstrap CI Methods for 
       different sample sizes on Gamma Dist.") +
  theme_light() +
  theme(legend.position="bottom")

@@

log_normal_fig <- ggplot(data = log_normal_df_long, 
                    aes(x = Sample_Size, y = prop, color = method)) + 
  geom_point() + geom_line() + 
  labs(title ="Prop. of CIs that Contain True Mean from Log-Normal Distibution",
       x = "Sample Size", 
       y = "Proportion (of 10,000 Simulations)", color = "Method",
       caption = "Fig. 10: Graph showing the Accuracy of Bootstrap CI Methods 
       for different sample sizes on Log-Normal Dist.") +
  theme_light() +
  theme(legend.position="bottom")

@@

```{r}
normal_fig
```

@@

```{r}
gamma_fig
```

@@

```{r}
log_normal_fig
```

>>>>

abstract: |
  This paper explores different bootstrap methods for creating confidence intervals and tests their performance against each other for different types of distributions. This paper will re-introduce the idea of bootstraps and confidence intervals, as well as the theory behind different methods of bootstrapped confidence intervals. Along with the theoretical details of each method, we will provide an example of each bootstrapped confidence interval method using a toy dataset. Then, we will run a simulation that evaluates the performance of a couple of the methods on a variety of sample sizes and types of distributions.

@@

We demonstrate performing a non-parametric bootstrap below using `SnowGR`, which gives the official snowfall dataset by the month for Grand Rapids, MI, starting in 1893 and is found in the `mosaic` package [@mosaic].
We will be using the `Dec` variable, which is the number of inches of snow that fell in December of each year. 
In particular, we are interested in the sample mean and later finding confidence intervals for the population mean of `Dec`. 
The distribution is shown (Figure 1).
```{r SnowGR histogram, fig.height = 4, fig.width = 5}
data(SnowGR)
gf_histogram(~ Dec, data = SnowGR) +
  labs(x = "Annual Snowfall in December", 
       title = "Distribution of Annual Snowfall in December", y = "Count",
       caption = "Fig. 1: Histogram of SnowGR")
```

Next, we will perform the non-parametric bootstrap using 10,000 replications.
```{r non-paramteric bootstrap}
set.seed(495)
orig_mean <- mean(~ Dec, data = SnowGR) # theta hat
nboot <- 10000
# initialize storage vectors
mean <- rep(0,nboot)
sd <- rep(0,nboot)

for (i in 1:nboot) {
  resampled <- as.data.frame(mosaic::resample(SnowGR, replace = TRUE))
  mean[i] <- mean(~Dec, data = resampled)
  sd[i] <- sd(~Dec, data = resampled)
}
dec_means <- as.data.frame(mean)
```
Figure 2 shows the histogram of bootstrapped values. Since we performed a large number of bootstrap replications, the bootstrapped distribution appears to be approximately Normal.
```{r, fig.height = 3, fig.width = 4}
gf_dhistogram(~ mean, data = dec_means) %>%
  gf_dens() %>%
  gf_labs(title = "Histogram of 10,000 Bootstrapped Mean Dec Values",
          caption = "Fig. 2: Bootstrap histogram of Dec values")
```

@@

```{r, fig.height = 4, fig.width = 5}
gf_histogram(~ Dec, data = SnowGR) +
  labs(x = "Annual Snowfall in December", 
       title = "Distribution of Annual Snowfall in December 
       with Standard Interval", y = "Count",
       caption = "Fig. 3: Histogram with Standard Interval Shown") +
  geom_vline(aes(xintercept= 13.85639)) +
  geom_vline(aes(xintercept= 17.65957))
```

@@

```{r}
qdata(~ mean, c(0.025, 0.975), data = dec_means)
```
Therefore, we can say that we are 95% confident that the true population mean of inches of snow that fall in December is between 13.91 inches and 17.72 inches. Our original histogram with the percentile confidence interval is shown (Figure 4).
```{r,fig.height = 4, fig.width = 5}
gf_histogram(~ Dec, data = SnowGR) +
  labs(x = "Annual Snowfall in December", 
       title = "Distribution of Annual Snowfall in December
       with Percentile Interval", y = "Count",
       caption = "Fig. 4: Histogram with Percentile Interval Shown") +
  geom_vline(aes(xintercept= 13.91170)) +
  geom_vline(aes(xintercept= 17.72353))
```

This interval is similar to the one created using the standard method, but shifted slightly to the right.

@@

```{r, fig.height = 4, fig.width = 5}
gf_histogram(~ Dec, data = SnowGR) +
  labs(x = "Annual Snowfall in December", 
       title = "Distribution of Annual Snowfall in December
       with BC interval", y = "Count",
       caption = "Fig. 5: Histogram with BC Interval Shown") +
  geom_vline(aes(xintercept= 14.01172)) +
  geom_vline(aes(xintercept= 17.82017))

@@

```{r, fig.height = 4, fig.width = 5}
gf_histogram(~ Dec, data = SnowGR) +
  labs(x = "Annual Snowfall in December", 
       title = "Distribution of Annual Snowfall in December
       with BCa interval", y = "Count",
       caption = "Fig. 6: Histogram with BCa Interval Shown") +
  geom_vline(aes(xintercept= 13.91591)) +
  geom_vline(aes(xintercept= 17.72353))
```

@@

```{r, fig.height = 4, fig.width = 5}
gf_histogram(~ Dec, data = SnowGR) +
  labs(x = "Annual Snowfall in December", 
       title = "Distribution of Annual Snowfall in December
       with Studentized Interval", y = "Count", 
       caption = "Fig. 7: Histogram with Studentized Interval Shown") +
  geom_vline(aes(xintercept= 14.03467)) +
  geom_vline(aes(xintercept= 18.01709)) 
```

@@

## Code
For our simulation, we will randomly sample from distributions in which we know the true population mean. 
For example, we will be sampling from a Gamma distribution where we can find the true population mean using the shape and rate parameters of the sample distribution. 
After we sample from the distribution, we will create a bootstrapped confidence interval for a few of the methods we described in the exposition. 
Since we know the true population mean of the distribution, we will determine if the true population mean is contained in the confidence interval for each of the methods we are testing [@Guillaume21]. 
We will perform this process a large number of times (10,000 simulations) and then average by the number of iterations to determine, on average, the proportion that each bootstrapped confidence interval method contains the true population mean for a given distribution. 
We will do this for a couple of different distributions and ranges of sample sizes to show how the intervals perform with skewed distributions and larger sample sizes.
We will define the accuracy of each method as the proportion that each bootstrapped confidence interval method contains the true population mean for a distribution

To perform this simulation, we will use the `boot` package [@boot]. 
This package can compute a bootstrap object which contains the output from a bootstrap calculation of a given sample and then produces confidence intervals from the bootstrap object. 
In particular, we will use the `boot.ci` function to create four confidence intervals: the standard (or normal) confidence interval, the studentized confidence interval, the percentile confidence interval, and the bias-corrected with acceleration confidence interval.

@@

```{r normal simulation}
run_sim_normal <- function(sim_reps) {
  # create function to return mean and sd of bootstrap replication
  boot.stat <- function(d, i) {
    d <- X[i]
    return (c(mean(d), sd(d)))
  }
  # initialize matrix to hold different bootstrap prop. 
  # for each sample size
  master <- matrix(nrow = num_sample_sizes, ncol = 5)
  # iterate through each sample size
  for(i in 1:num_sample_sizes) {
    # initialize vectors that hold if the CI contains true mean 
    # by iteration
    norm_contains <- rep(0, sim_reps)
    student_contains <- rep(0, sim_reps)
    percentile_contains <- rep(0, sim_reps)
    bca_contains <- rep(0, sim_reps)
    # 
    n <- sample_sizes[i]
    true_mean <- 0
    # perform 10,000 iterations of bootstraps
    for (j in 1:sim_reps) {
      X <- rnorm(n, 0, 1)
      values <- data.frame(X)
      # perform bootstrap and CIs
      results <- boot(data = values, statistic = boot.stat, R = boot_reps)
      ci <- boot.ci(results, conf = 0.95, 
                    type = c("norm", "perc", "bca", "stud"))
      # update vectors depending on if the CI contains true mean
      norm_contains[j] <- true_mean >= ci$normal[2] & 
        true_mean <= ci$normal[3]
      student_contains[j] <- true_mean >= ci$student[4] & 
        true_mean <= ci$student[5]
      percentile_contains[j] <- true_mean >= ci$percent[4] & 
        true_mean <= ci$percent[5]
      bca_contains[j] <- true_mean >= ci$bca[4] &
        true_mean <= ci$bca[5]
    }
    # add the sample size and average proportion to matrix
    master[i, 1] <- sample_sizes[i]
    master[i, 2] <- sum(norm_contains)/sim_reps
    master[i, 3] <- sum(student_contains)/sim_reps
    master[i, 4] <- sum(percentile_contains)/sim_reps
    master[i, 5] <- sum(bca_contains)/sim_reps
  }
  # some wrangling
  master_df <- as.data.frame(master) %>% 
      rename("Sample_Size" = V1, "Normal" = V2, "Studentized" = V3, 
             "Percentile" = V4, "BCa" = V5) 
  master_df
}
```

@@

normal_fig <- ggplot(data = normal_df_long, 
                    aes(x = Sample_Size, y = prop, color = method)) + 
  geom_point() + geom_line() + 
  labs(
    title = "Accuracy of Bootstrap CIs using Normal Distibution", 
    x = "Sample Size", 
    y = "Proportion (of 10,000 Simulations)", color = "Method",
    caption = "Fig. 8: Graph showing the Accuracy of Bootstrap CI 
    Methods for different sample sizes on Normal Distribution") +
  theme_light() +
  theme(legend.position="bottom")

@@

gamma_fig <- ggplot(data = gamma_df_long, 
                    aes(x = Sample_Size, y = prop, color = method)) + 
  geom_point() + geom_line() + 
  labs(title = "Accuracy of Bootstrap CIs using Gamma Distibution", 
       x = "Sample Size", 
       y = "Proportion (of 10,000 Simulations)", color = "Method",
       caption = "Fig. 9: Graph showing the Accuracy of Bootstrap CI 
       Methods for different sample sizes on Gamma Distribution") +
  theme_light() +
  theme(legend.position="bottom")

@@

log_normal_fig <- ggplot(data = log_normal_df_long, 
                    aes(x = Sample_Size, y = prop, color = method)) + 
  geom_point() + geom_line() + 
  labs(title ="Accuracy of Bootstrap CIs using Log-Normal 
       Distibution",
       x = "Sample Size", 
       y = "Proportion (of 10,000 Simulations)", color = "Method",
       caption = "Fig. 10: Graph showing the Accuracy of Bootstrap CI 
       Methods for different sample sizes on Log-Normal Distribution") +
  theme_light() +
  theme(legend.position="bottom")

@@

```{r,fig.height = 4, fig.width = 5}
normal_fig
```

@@

```{r, fig.height = 4, fig.width = 5}
gamma_fig
```

@@

```{r, fig.height = 4, fig.width = 5}
log_normal_fig
```
